Evolution Reinforces Cooperation with the Emergence of Self-Recognition Mechanisms: an Empirical Study of Strategies in the Moran process for the Iterated Prisoner's Dilemma

### 2018-08-10 tf1.tex
Re rename all images.

% Template for PLoS
% Version 3.4 January 2017
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended
% to minimize problems and delays during our production
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% Once your paper is accepted for publication,
% PLEASE REMOVE ALL TRACKED CHANGES in this file
% and leave only the final text of your manuscript.
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file.
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission.
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column.
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2".
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace}
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

% Header and Footer with logo
\usepackage{lastpage,fancyhdr}
\usepackage{graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{27.023pt}
\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

%% END MACROS SECTION

\usepackage{algorithm,algorithmic}
\usepackage{amsmath}
\usepackage{authblk}
\usepackage{booktabs}
\usepackage{booktabs}
\usepackage{fancyvrb}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{multicol}
\usepackage{pdflscape}
\usepackage{standalone}
\usepackage{subcaption}
\usepackage{tikz}
\usepackage{xstring}

\usetikzlibrary{calc, shapes, patterns}

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline
{Evolution reinforces cooperation with the emergence of self-recognition
 mechanisms: an empirical study of strategies in the Moran process for the
 iterated prisoner's dilemma }
% Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Vincent Knight\textsuperscript{1*\Yinyang},
Marc Harper\textsuperscript{2\Yinyang},
Nikoleta E. Glynatsi\textsuperscript{2\ddag},
Owen Campbell\textsuperscript{3\ddag},
\\
\bigskip
\textbf{1} Cardiff University, School of Mathematics, Cardiff, UK
\\
\textbf{2} Google Inc., Mountain View, CA, USA
\\
\textbf{3} Independent Researcher, Chester, UK
\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
%
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
\Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
\ddag These authors also contributed equally to this work.

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* knightva@cardiff.ac.uk

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
    We present insights and empirical results from an extensive numerical
    study of the evolutionary dynamics of the iterated prisoner's dilemma.
    Fixation probabilities for Moran processes are obtained
    for all pairs of 164 different strategies including classics such as TitForTat, zero
    determinant strategies, and many more sophisticated strategies.
    Players with long memories and sophisticated behaviours outperform
    many strategies that perform well in a two player setting. Moreover we
    introduce several strategies trained with evolutionary algorithms to
    excel at the Moran process. These strategies are excellent invaders
    and resistors of invasion and in some cases naturally evolve handshaking
    mechanisms to resist invasion. The best invaders were those trained
    to maximize total payoff while the best resistors invoke handshake mechanisms.
    This suggests that while maximizing individual payoff
    can lead to the evolution of cooperation through invasion, the relatively
    weak invasion resistance of payoff maximizing strategies are not as
    evolutionarily stable as strategies employing handshake mechanisms.
\linenumbers

\section*{Introduction}

The Prisoner's Dilemma (PD) \cite{Flood1958} is a fundamental two player game
used to model a variety of strategic interactions. Each player chooses
simultaneously and independently
between cooperation (C) or defection (D). The payoffs of
the game are defined by the matrix $\begin{pmatrix} R & S \\ T & P
\end{pmatrix}$, where $T > R > P > S$ and $2R > T + S$. The PD is a one
round game, but is commonly studied in a manner where the prior outcomes
matter. This repeated form is called the Iterated Prisoner's
Dilemma (IPD). As described in \cite{Axelrod1980a, Knight2016, Press2012} a
number of strategies have been developed to take advantage of the history of
play. Strategies referred to as zero determinant (ZD) strategies
\cite{Press2012} can manipulate some players through extortionate mechanisms.

The Moran Process \cite{Moran1957} is a model of
evolutionary population dynamics that has been used to gain insights about the
evolutionary stability in a number of settings.
Several earlier
works have studied iterated games in the context of the prisoner's dilemma
\cite{Nowak, stewart2013extortion}, however these often make simplifying assumptions
or are limited to classes of strategies
such as memory-one strategies that only use the previous round of play.

This manuscript provides a detailed numerical analysis of agent-based simulations
of \textbf{164} complex and adaptive strategies for the
IPD\@. This is made possible by the Axelrod library \cite{axelrodproject}, an
effort to provide software for reproducible research for the IPD\@. The library
now contains over 186 parameterized
strategies including classics like TitForTat and WinStayLoseShift, as well as
recent variants such as OmegaTFT, zero determinant and other memory one
strategies, strategies based on finite state machines, lookup tables, neural
networks, and other machine learning based strategies, and a collection of novel
strategies. Not all strategies have been considered for this study: excluded
are those that make use of knowledge of the number of turns in a match
and others that have a high
computational run time. The large number of strategies are available thanks to
the open source nature of the project with over 50 contributors from around the
world, made by programmers and researchers \cite{Knight2016}. Three of the considered
strategies are finite state machines trained specifically for Moran processes
(described further in the Methods section).

In addition to providing a large collection of strategies, the Axelrod library
can conduct matches, tournaments and population
dynamics with variations including noise and spatial structure.
The strategies and simulation frameworks are
automatically tested to an extraordinarily high degree of coverage in accordance
with best research software practices.

Using the Axelrod library and the many strategies it contains, we obtain the
probability with which a given strategy takes over a population (referred to as
fixation probability) for all pairs of strategies, identifying those that are
effective invaders and those resistant to invasion, for population sizes $N=2$
to $N=14$. Moreover we present a number (16) of strategies that were created via
reinforcement algorithms (evolutionary and particle swarm algorithms) that are
among the best invaders and resistors of invasion known to date, and show that
handshaking mechanisms naturally arise from these processes as an
invasion-resistance mechanism.

In 2016, work has argued that agent-based
simulations can provide insights in evolutionary game theory not available
via direct mathematical analysis~\cite{adami2016evolutionary}. The results
and insights contained in this paper would be difficult to derive analytically.

In particular the following questions are addressed:
\begin{enumerate}
    \item What strategies are good invaders?
    \item What strategies are good at resisting invasion?
    \item How does the population size affect these findings?
\end{enumerate}

While the results agree with some of the published literature, it is found that:

\begin{enumerate}
 \item Zero determinant strategies are not effective invaders or defenders for $N > 2$
 \item Complex strategies can be effective, and in fact can naturally evolve
     through evolutionary processes to outperform designed strategies.
 \item The strongest resistors specifically evolve or possess a handshake mechanism.
 \item Strong invaders are generally cooperative strategies that do not defect
 first but retaliate to varying degrees of intensity against strategies that defect.
 \item Strategies evolved to maximize their total payoff can be strong invaders
 and achieve mutual cooperation with many other strategies.
\end{enumerate}

\section*{Materials and methods}
\subsection*{The Moran Process}

A Moran process is a stochastic birth death process on a finite population in which the
population size stays constant over time. Individuals are \textit{selected}
according to a given fitness landscape. The fitness landscape in this work is
defined as the total utility against all other individuals in the population.
Once selected, the individual
is reproduced and similarly another individual is chosen to be removed from the
population (a uniform random selection is used). This is shown diagrammatically
in Fig~\ref{fig:moran_process}. In some settings mutation is
also considered but without mutation (the case considered in this work) this
process will arrive at an absorbing state where the population is entirely made
up of players of one strategy. The probability with which a given strategy,
starting from a single individual takes
over a population is called the \textit{fixation probability}. A more detailed
analytic description of this is given later. In our
simulations offspring do not inherit any knowledge or history from parent
replicants.

\begin{figure}[!hbtp]
    \centering
    \scalebox{.5}{

\begin{tikzpicture}[scale=.9]
	\node (A1) at (-1, -1) [circle, pattern=north west lines, pattern
        color=blue!70, draw=blue] {};
	\node (A2) at (-1, 1) [circle, pattern=north west lines, pattern
        color=blue!70, draw=blue] {};
	\node (A3) at (0, .2) [circle, pattern=north west lines, pattern
        color=blue!70, draw=blue] {};
	\node (A4) at (.2, -.6) [circle, pattern=north west lines, pattern
        color=blue!70, draw=blue] {};
	\node (A5) at (.5, -0.1) [circle, pattern=north west lines, pattern
        color=blue!70, draw=blue] {};
	\node (B1) at (-1, -.2) [circle, pattern=north east lines, pattern
        color=red!70, draw=red] {};
	\node (B2) at (1, -.5) [circle, pattern=north east lines, pattern
        color=red!70, draw=red] {};
	\node (B3) at (.5, .8) [circle, pattern=north east lines, pattern
        color=red!70, draw=red] {};
	\node (B4) at (-.6, .2) [circle, pattern=north east lines, pattern
        color=red!70, draw=red] {};
	\node (B5) at (.6, -.8) [circle, pattern=north east lines, pattern
        color=red!70, draw=red] {};

	\draw [->] (1, 0) -- (3, 0) node [above, pos=0.5] {Selection};

	\node (A1) at ($(A1) + (4.5, 0)$) [circle, pattern=north west lines,
        pattern color=blue!70, draw=blue] {};
	\node (A2) at ($(A2) + (4.5, 0)$) [circle, pattern=north west lines,
        pattern color=blue!70, draw=blue] {};
	\node (A3) at ($(A3) + (4.5, 0)$) [circle, pattern=north west lines,
        pattern color=blue!70, draw=blue] {};
	\node (A4) at ($(A4) + (4.5, 0)$) [circle, pattern=north west lines,
        pattern color=blue!70, draw=blue] {};
    \node (A5) at ($(A5) + (4.5, 0)$) [circle, pattern=north west lines,
        pattern color=blue!70, draw=blue] {};
	\node (B1) at ($(B1) + (4.5, 0)$) [circle, pattern=north east lines,
        pattern color=red!70, draw=red] {};
	\node (B2) at ($(B2) + (4.5, 0)$) [circle, pattern=north east lines,
        pattern color=red!70, draw=red] {};
	\node (B3) at ($(B3) + (4.5, 0)$) [circle, pattern=north east lines,
        pattern color=red!70, draw=red] {};
	\node (B4) at ($(B4) + (4.5, 0)$) [circle, pattern=north east lines,
        pattern color=red!70, draw=red] {};
	\node (B5) at ($(B5) + (4.5, 0)$) [circle, pattern=north east lines,
        pattern color=red!70, draw=red] {};

	\draw [dotted, very thick] (B4) circle (.25cm);

	\draw [->] (5.5, 0) -- (7.5, 0) node [above, pos=0.5] {Birth};

	\node (A1) at ($(A1) + (4.5, 0)$) [circle, pattern=north west lines,
        pattern color=blue!70, draw=blue] {};
	\node (A2) at ($(A2) + (4.5, 0)$) [circle, pattern=north west lines,
        pattern color=blue!70, draw=blue] {};
	\node (A3) at ($(A3) + (4.5, 0)$) [circle, pattern=north west lines,
        pattern color=blue!70, draw=blue] {};
	\node (A4) at ($(A4) + (4.5, 0)$) [circle, pattern=north west lines,
        pattern color=blue!70, draw=blue] {};
    \node (A5) at ($(A5) + (4.5, 0)$) [circle, pattern=north west lines,
        pattern color=blue!70, draw=blue] {};
	\node (B1) at ($(B1) + (4.5, 0)$) [circle, pattern=north east lines,
        pattern color=red!70, draw=red] {};
	\node (B2) at ($(B2) + (4.5, 0)$) [circle, pattern=north east lines,
        pattern color=red!70, draw=red] {};
	\node (B3) at ($(B3) + (4.5, 0)$) [circle, pattern=north east lines,
        pattern color=red!70, draw=red] {};
	\node (B4) at ($(B4) + (4.5, 0)$) [circle, pattern=north east lines,
        pattern color=red!70, draw=red] {};
	\node (B5) at ($(B5) + (4.5, 0)$) [circle, pattern=north east lines,
        pattern color=red!70, draw=red] {};

	\draw [dotted, very thick] (B4) circle (.25cm);
	\node (B6) at ($(B4) + (0.5, 0.5)$) [circle, pattern=north east lines,
        pattern color=red!70, draw=red] {};
	\draw [->, dotted, very thick] (B4) [out=90, in=180] to (B6);

	\draw [->] (10, 0) -- (12, 0) node [above, pos=0.5] {Selection};

	\node (A1) at ($(A1) + (4.5, 0)$) [circle, pattern=north west lines,
        pattern color=blue!70, draw=blue] {};
	\node (A2) at ($(A2) + (4.5, 0)$) [circle, pattern=north west lines,
        pattern color=blue!70, draw=blue] {};
	\node (A3) at ($(A3) + (4.5, 0)$) [circle, pattern=north west lines,
        pattern color=blue!70, draw=blue] {};
	\node (A4) at ($(A4) + (4.5, 0)$) [circle, pattern=north west lines,
        pattern color=blue!70, draw=blue] {};
    \node (A5) at ($(A5) + (4.5, 0)$) [circle, pattern=north west lines,
        pattern color=blue!70, draw=blue] {};
	\node (B1) at ($(B1) + (4.5, 0)$) [circle, pattern=north east lines,
        pattern color=red!70, draw=red] {};
	\node (B2) at ($(B2) + (4.5, 0)$) [circle, pattern=north east lines,
        pattern color=red!70, draw=red] {};
	\node (B3) at ($(B3) + (4.5, 0)$) [circle, pattern=north east lines,
        pattern color=red!70, draw=red] {};
	\node (B4) at ($(B4) + (4.5, 0)$) [circle, pattern=north east lines,
        pattern color=red!70, draw=red] {};
	\node (B5) at ($(B5) + (4.5, 0)$) [circle, pattern=north east lines,
        pattern color=red!70, draw=red] {};
	\node (B6) at ($(B6) + (4.5, 0)$) [circle, pattern=north east lines,
        pattern color=red!70, draw=red] {};

	\draw [dotted, very thick] (A3) circle (.25cm);

	\draw [->] (14.5, 0) -- (16.5, 0) node [above, pos=0.5] {Death};

	\node (A1) at ($(A1) + (4.5, 0)$) [circle, pattern=north west lines,
        pattern color=blue!70, draw=blue] {};
	\node (A2) at ($(A2) + (4.5, 0)$) [circle, pattern=north west lines,
        pattern color=blue!70, draw=blue] {};
	\node (A3) at ($(A3) + (4.5, 0)$) {};
	\node (A4) at ($(A4) + (4.5, 0)$) [circle, pattern=north west lines,
        pattern color=blue!70, draw=blue] {};
    \node (A5) at ($(A5) + (4.5, 0)$) [circle, pattern=north west lines,
        pattern color=blue!70, draw=blue] {};
	\node (B1) at ($(B1) + (4.5, 0)$) [circle, pattern=north east lines,
        pattern color=red!70, draw=red] {};
	\node (B2) at ($(B2) + (4.5, 0)$) [circle, pattern=north east lines,
        pattern color=red!70, draw=red] {};
	\node (B3) at ($(B3) + (4.5, 0)$) [circle, pattern=north east lines,
        pattern color=red!70, draw=red] {};
	\node (B4) at ($(B4) + (4.5, 0)$) [circle, pattern=north east lines,
        pattern color=red!70, draw=red] {};
	\node (B5) at ($(B5) + (4.5, 0)$) [circle, pattern=north east lines,
        pattern color=red!70, draw=red] {};
	\node (B6) at ($(B6) + (4.5, 0)$) [circle, pattern=north east lines,
        pattern color=red!70, draw=red] {};

	\draw [dotted, very thick] (A3) circle (.25cm);
\end{tikzpicture}

}
    \caption{A diagrammatic representation of a Moran process.}
    \label{fig:moran_process}
\end{figure}

The Moran process was initially introduced in \cite{Moran1957}. It has since
been used in a variety of settings including the understanding of the spread of
cooperative and non-cooperative behaviour such as cancer \cite{West2016} and the
emergence of cooperative behaviour in spatial topologies \cite{Nowak2017}.
However these works mainly consider relatively simple strategies. A few works
looked at evolutionary stability of agent-based strategies within the Prisoner's Dilemma
\cite{Li2014} but this is not done in the more widely used setting of the Moran
process, rather in terms of infinite population stability. In \cite{Baek2016}
Moran processes are studied in a theoretical framework for a small subset of
strategies. The subset included memory one strategies: strategies that recall
the events of the previous round only.

Of particular interest are the zero determinant strategies introduced in
\cite{Press2012}. It was argued in \cite{stewart2013extortion} that generous
ZD strategies are robust against invading strategies. However, in \cite{Lee2015},
a strategy using machine learning techniques was capable of resisting invasion
and also able to invade any memory one strategy. In 2017, \cite{Hilbe2017}
has investigated the effect of memory length on strategy performance and the
emergence of cooperation but this is not done in a Moran process context and only
considers specific cases of memory 2 strategies. In \cite{Adami2013} it was
recognised that many zero determinant strategies do not fare well against
themselves. This is a disadvantage for the Moran process where the best
strategies cooperate well with other players using the same strategy.

This work uses pair-wise Moran processes in a similar way to matches in the many
IPD tournaments published since Axelrod's original work~\cite{Axelrod1980a}.
A population-based perspective is given which adds additional
evolutionary components to the IPD, namely the evolutionary dynamics of invasion
and resistance.

\subsection*{Strategies considered}\label{sec:strategies}

To carry out this numerical experiment, 164
strategies, listed (with their properties) in the Appendix,
are used from the Axelrod library. There are
43 stochastic and
121 deterministic strategies. Their memory depth,
defined by the number of rounds of history used by the strategy each round, is
shown in Table~\ref{tbl:memory_depth_count}. The memory depth is infinite if the
strategy uses the entire history of play (whatever its length). For example, a
strategy that utilizes a handshaking mechanism where the opponent's actions on
the first few rounds of play determines the strategies subsequent behavior would
have infinite memory depth.

A number of these strategies have been trained with reinforcement learning
algorithms prior to this study and not specifically for the Moran process.

\begin{itemize}
    \item Evolved ANN: a neural network based strategy;
    \item Evolved LookerUp: a lookup table based strategy;
    \item PSO Gambler: a stochastic version of the lookup table based strategy;
    \item Evolved HMM: a hidden Markov model based strategy.
\end{itemize}

Apart from the PSO Gambler strategy, which was trained using a particle swarm
optimisation algorithm, these strategies are trained with an evolutionary
algorithm that perturbs strategy parameters and optimizes the mean total score
against all other opponents \cite{affenzeller2009genetic}. They were trained to
win IPD tournaments by maximizing their mean total payoffs against a variety
of opponents. Variation is
introduced via mutation and crossover of parameters, and the best performing
strategies are carried to the next generation along with new variants. Similar
methods appear in the literature \cite{Ashlock2006}.
There has also been some work on strategies using an evolutionary algorithm in real time:
in~\cite{Gaudesi2016}
an evolutionary algorithm is used to build a model of the opponent and attempt
to exploit any potential weakness. In this work all strategies resulting from
evolutionary algorithms are pre-trained.

More information about each player can be obtained in the documentation for
\cite{axelrodproject} and a detailed description of the performance
of these strategies in IPD tournaments is described in~\cite{Harper2017}.

All of the training code is archived at \cite{marc_harper_2017_824264}. This
software is (similarly to the Axelrod library) available on
github\footnote{\url{https://github.com/Axelrod-Python/axelrod-dojo}}
with documentation to
train new strategies easily. Training
typically takes less than 100 generations and can be completed within several
hours on commodity hardware.

There are three further strategies trained specifically for this study; Trained
FSM 1, 2, and 3 (TF1 - TF3). These are based on finite state machines of 16, 16,
and 8 states respectively. These are shown in Fig~\ref{fig:tf1},~\ref{fig:tf2}
and~\ref{fig:tf3}), using the notation common in the literature where
\(A_1/A_2\) is the action of the opponent \(A_1\) and the response of the player
\(A_2\) as well as arrows corresponding to changes of state.

\begin{figure}[!hbtp]
    \centering
    \scalebox{.7}{
    \begin{tikzpicture}

@@

    \end{tikzpicture}
    }
    \caption{TF1: a 16 state finite state machine with a handshake leading to
    mutual cooperation at state 4.}
    \label{fig:tf1}
\end{figure}

\begin{figure}[!hbtp]
    \centering
    \scalebox{.7}{

    \begin{tikzpicture}

    \tikzstyle{state}=[minimum width=0.8cm, font=\boldmath];
    \node[circle, draw=black, thick] (1) at (0, 0) [state] {$2$};
	\node[circle, draw=black, thick] (4) at (4, 0) [state] {$5$};

	\node[circle, draw=black, thick] (5) at (-2, -1) [state] {$6$};
    \node[circle, draw=black, thick] (6) at (2, -1) [state] {$7$};
    \node[circle, draw=black, thick] (3) at ($(5)+(-1,-2)$) [state] {$4$};
    \node[circle, draw=black, thick] (12) at ($(5)+(0,-4)$) [state] {$13$};
    \node[circle, draw=black, thick] (8) at ($(12)+(0,-2)$) [state] {$9$};
    \node[circle, draw=black, thick] (9) at ($(8)+(-2,0)$) [state] {$10$};
    \node[circle, draw=black, thick] (0) at ($(8)+(0,-2)$) [state] {$1$};

    \node[circle, draw=black, thick] (2) at ($(6)+(0,-2)$) [state] {$3$};
    \node[circle, draw=black, thick] (14) at ($(2)+(2,0)$) [state] {$15$};
    \node[circle, draw=black, thick] (7) at ($(2)+(0,-2)$) [state] {$8$};
    \node[circle, draw=black, thick] (10) at ($(14)+(0,-2)$) [state] {$11$};
    \node[circle, draw=black, thick] (13) at ($(7)+(0,-2)$) [state] {$14$};
    \node[circle, draw=black, thick] (11) at ($(10)+(0,-2)$) [state] {$12$};
    \node[circle, draw=black, thick] (15) at ($(11)+(0,-2)$) [state] {$16$};

    \coordinate[below of=0] (s);

    \draw (s) edge[out=90, in=-90, ->, thick] node [left] {$C$} (0);

    \draw (0) edge[out=0, in=-90, ->, thick] node [below right] {$C/D$} (13);
    \draw (0) edge[out=45, in=0, ->, thick] node [above, rotate=90] {$D/D$} (12);

    \draw (1) edge[out=-90, in=0, ->, thick] node [above, yshift=1.3cm, xshift=0.3cm] {$C/D$} (3);
    \draw (1) edge[out=0, in=180, ->, thick] node [above] {$D/D$} (4);

    \draw (2) edge[out=-45, in=-135, ->, thick] node [above] {$C/D$} (14);
    \draw (2) edge[out=180, in=90, ->, thick] node [above, yshift=0.8cm, xshift=3cm] {$D/D$} (9);

    \draw (3) edge[out=180, in=-135, ->, thick] node [left, yshift=2cm, xshift=-0.1cm] {$C/C$} (0);
    \draw (3) edge[out=135, in=180, ->, thick] node [above, yshift=0.2cm] {$D/D$} (1);

    \draw (4) edge[out=90, in=90, ->, thick] node [above] {$C/D$} (1);
    \draw (4) edge[out=-90, in=0, ->, thick] node [right] {$D/D$} (2);

    \draw (5) edge[out=-90, in=90, ->, thick] node [right, yshift=1cm] {$C/C$} (12);
    \draw (5) edge[out=0, in=180, ->, thick] node [below right] {$D/C$} (6);

    \draw (6) edge[out=135, in=-45, ->, thick] node [right, xshift=1cm] {$C/C$} (1);
    \draw (6) edge[out=-45, in=135, ->, thick] node [above left, yshift=-0.5cm] {$D/D$} (14);

    \draw (7) edge[out=135, in=45, ->, thick] node [above] {$C/D$} (12);
    \draw (7) edge[out=90, in=-90, ->, thick] node [left] {$D/D$} (2);

    \draw (8) edge[out=90, in=-135, ->, thick] node [above right] {$C/D$} (7);
    \draw (8) edge[out=180, in=0, ->, thick] node [above] {$D/D$} (9);

    \draw (9) edge[out=-45, in=-135, ->, thick] node [below] {$C/D$} (8);
    \draw (9) edge[out=-90, in=180, ->, thick] node [right] {$D/D$} (0);

    \draw (10) edge[out=180, in=-60, ->, thick] node [below, yshift=-0.5cm, xshift=0.4cm] {$C/C$} (2);
    \draw (10) edge[out=0, in=0, ->, thick] node [right] {$D/C$} (15);

    \draw (11) edge[out=90, in=-45, ->, thick] node [below right, xshift=0.7cm] {$C/D$} (7);
    \draw (11) edge[out=180, in=0, ->, thick] node [below] {$D/D$} (13);

    \draw (12) edge[out=135, in=-45, ->, thick] node [above left] {$C/C$} (3);
    \draw (12) edge[out=-90, in=135, ->, thick] node [left] {$D/D$} (8);

    \draw (13) edge[out=90, in=-90, ->, thick] node [left] {$C/C$} (7);
    \draw (13) edge[out=45, in=-135, ->, thick] node [below, yshift=-0.2cm, rotate=45] {$D/D$} (10);

    \draw (14) edge[out=-90, in=90, ->, thick] node [right] {$C/D$} (10);
    \draw (14) edge[out=-100, in=45, ->, thick] node [above] {$D/D$} (7);

    \draw (15) edge[out=-135, in=180, ->, thick, loop] node [left] {$C/C$} (15);
    \draw (15) edge[out=90, in=-90, ->, thick] node [left] {$D/D$} (11);

    \end{tikzpicture}
    }
    \caption{TF2: a 16 state finite state machine with a handshake leading to
    mutual cooperation at state 16.}
    \label{fig:tf2}
\end{figure}

\begin{figure}[!hbtp]
    \centering
    \begin{tikzpicture}

    \tikzstyle{state}=[minimum width=0.5cm, font=\boldmath];

    \node[circle, draw=black, thick]  (0) at (0,0) [state] {$1$};
    \node[circle, draw=black, thick]  (1) at ($(0)+(0,2)$) [state] {$2$};
    \node[circle, draw=black, thick]  (2) at ($(1)+(2,1.5)$) [state] {$3$}; 
    \node[circle, draw=black, thick]  (3) at ($(2)+(3,0)$) [state] {$4$};
    \node[circle, draw=black, thick]  (4) at ($(1)+(8,0)$) [state] {$5$};  
    \node[circle, draw=black, thick]  (5) at ($(0)+(8,0)$) [state] {$6$};
    \node[circle, draw=black, thick]  (6) at ($(3)+(0,-4.5)$) [state] {$7$};
    \node[circle, draw=black, thick]  (7) at ($(2)+(0,-4.5)$) [state] {$8$};

    \draw (s) edge[out=0, in=180, ->, thick] node [above] {$C$} (0);

    \draw (0) edge[out=-45, in=-100, loop, thick] node [below] {$C/C$} (0);
    \draw (6) edge[out=-45, in=-100, loop, thick] node [below] {$C/D$} (6);
    \draw (6) edge[out=190, in=135, loop, thick] node [below, yshift=-0.5cm] {$D/D$} (6);
    \draw (7) edge[out=190, in=135, loop, thick] node [above, xshift=1cm] {$C/D$} (7);
    \draw (2) edge[out=190, in=135, loop, thick] node [left] {$D/D$} (2);

    \draw (2) edge[out=0,in=180,->,thick] node [above] {$C/C$} (3);
    \draw (3) edge[out=-35,in=170,->,thick] node [above right] {$C/D$} (4);
    \draw (5) edge[out=-135,in=0,->,thick] node [below] {$C/C$} (6);

    \draw (0) edge[out=45,in=-135,->,thick] node [above, rotate=45, xshift=2cm, yshift=-0.5cm] {$D/C$} (3);

    \draw (1) edge[out=-45,in=180,->,thick] node [below right, xshift=2cm] {$C/D$} (5);
    \draw (1) edge[out=-135,in=135,->,thick] node [left] {$D/C$} (0);
    \draw (3) edge[out=-90,in=90,->,thick] node [above, rotate=90] {$D/D$} (6);
    \draw (4) edge[out=90,in=0,->,thick] node [above] {$C/C$} (3);
    \draw (4) edge[out=180,in=0,->,thick] node [above left, xshift=-2.5cm] {$D/D$} (1);
    \draw (5) edge[out=135,in=-45,->,thick] node [below, rotate=-45] {$D/D$} (3);
    \draw (7) edge[out=-90,in=-90,->,thick] node [below] {$D/C$} (5);
    \end{tikzpicture}
    \caption{TF3: an 8 state finite state machine.}
    \label{fig:tf3}
\end{figure}

As opposed to the previously described strategies, these strategies were trained
with the objective function of \textit{mean fixation probabilities for Moran
processes} starting at initial population states consisting of \(N/2\)
individuals of the training candidates and \(N/2\) individuals of an opponent
strategy, taken from a selection of 150 opponents from the Axelrod library:

\begin{itemize}
	\item TF1 \(N=12\), 0\% noise.
	\item TF2 \(N=10\), 0\% noise.
	\item TF3 \(N=8\), 1\% noise.
\end{itemize}

The trained algorithms
were run for fewer than 50 generations. Training data for this is available at
\cite{data}.

TF1 has an initial handshake of CCD and cooperates if the opponent matches.
However if the opponent later defects, TF1 will respond in kind, so the
handshake is not permanent. Only one player (Prober 4 \cite{Prison1998}) manages to
achieve cooperation with TF1 after about 20 rounds of play. TF1 is functionally
very similar to a strategy known as ``Collective Strategy'', which has a
handshake of CD and cooperates with opponents that matched the handshake
until they defect, defecting thereafter if the opponent ever defects \cite{Li2009}.
Collective Strategy was specifically designed for evolutionary processes.

>>>>

\documentclass[crop,tikz]{standalone}
\usepackage{tikz}
\usetikzlibrary{calc, shapes, patterns}
\begin{document}
    \begin{tikzpicture}

@@

    \end{tikzpicture}
\end{document}
